---
title: "ChemSpace"
author: "Daniel Martinez"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
  html_document:
    toc: true
    toc_float: true
    toc_depth: 2

knit: (function(input_file, encoding) {
  out_dir <- 'docs';
  rmarkdown::render(input_file,
 encoding=encoding,
 output_file=file.path(dirname(input_file), out_dir, 'index.html'))})
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, cache = F)
```

```{r message=FALSE, warning=FALSE}
library(tidyverse) # master library to deal with data frames
library(readxl) # read xlsx or xls files
library(ggrepel) # ggplot add-on, to plot names that don't collapse in same position
library(FactoMineR) # for PCA
library(factoextra) # for PCA
library(here) # usefull to save plots in folders given a root
library(viridis) # color palette package
library(plotly)
```



# Some tests with 3D PCA plots and drugbank database

## Data preparation and testing

I've downloaded the files from the Zimmermann paper from Nature, 2019. They have a big database of 2099 compounds from drugbank to which they run a Python script to extract the *chemical fingerprints*, using them to represent the chemical space via a PCA. 

```{r message=FALSE, warning=FALSE}
# biolog metabolites
biolog = read_csv('Biolog_metabolites.csv')

# drugbank coordinates
drugbank.coord = read.delim('.\\Data_drug_bacteria_gene_mapping\\Input\\drugbank_pca.txt', header = FALSE)
drugbank = read_csv('.\\Data_drug_bacteria_gene_mapping\\Input\\drugbank_approved_MW_150_1000_functional_groups_all.csv') %>% select(-Number_of_carboxylic_acids_1)

# plates to use from Biolog
plates = c("PM11C", "PM12B", "PM13B", "PM14A", "PM15B", "PM16A", "PM17A", "PM18C", "PM19" , "PM20B")
biolog.drug = biolog %>% filter(Plate %in% plates)
# list of drugs present in biolog plates
drugs = biolog.drug %>% select(Metabolite) %>% unique %>% t %>% as.character
```

Just a glimpse of the drug names:

```{r}
head(drugs)
```

And a glimpse of the huge data table from the paper:

```{r}
head(drugbank)
```

Doing some data transformations (including the names of biolog compounds and that into the huge data table), we notice one important thing: only ~70 compounds from biolog are in the original data table. Anyway, it's a good practice to see if our drugs are *occupying* enough of the drug space.

```{r}

drugbank.coord['Drugbank'] = drugbank$GENERIC_NAME

# create a new variable, just in case
test2 = drugbank.coord
# data transformation, create Category with Biolog (if the compound is Drugbank) and Drugbank (if not)
test2 = test2 %>%
  mutate(biolog = ifelse(Drugbank %in% drugs, Drugbank, NA),
         Category = ifelse(Drugbank %in% drugs, 'Biolog', 'Drugbank'),
         Category = as.factor(Category)) 

rownames(test2) = test2$Drugbank

head(test2)
```

And let's plot it in 3D with Plotly:

```{r}
fig = plot_ly(data = test2, x = ~V1, y = ~V2, z = ~V3, text = rownames(test2), 
               color = ~Category, colors = c('#FF0900', '#B8B8B8'),
               marker = list(size = 3))
fig

```

## New biolog drugs and PCA

In order to try to complete our dataset with more drugs, I've extracted the KEGG IDs from the biolog table, passed them into a website that converts them into PubChem IDs, and got the chemical fingerprints using an adapted version of the Python script available at their GitHub repository.  
This way, I've been able to get over a **100 new compounds** from biolog that were not in the original data table, but there are still ~70 compounds that need to be added. And also **metformin**!! Probably will be a good idea to put a label to metformin to see where it falls. 

```{r message=FALSE, warning=FALSE}
# first, separate KEGG ids to generate a new column to the biolog data frame with pubchem ids
kegg_ids = biolog %>%
  select(KEGG_ID) %>% t %>% as.character %>% unique 
# remove NAs
kegg_ids = kegg_ids[complete.cases(kegg_ids)]
# write list of genes
write.table(kegg_ids, 'KEGG_IDs_biolog.txt', quote = F, col.names = F, row.names = F)
## go here to convert: http://csbg.cnb.csic.es/mbrole/conversion.jsp

# read generated list and merge with biolog
kegg2pub =  read.csv("D:/MRC_Postdoc/Pangenomic/Chem_space/KEGG2PubChemIDs.txt")
biolog = biolog %>% left_join(kegg2pub)


### let's get the missing drugs (as many as we can, at least)
missing = drugs[!drugs %in% unique(drugbank$GENERIC_NAME)]

drugs_missing = biolog %>%
  filter(Metabolite %in% missing) %>%
  select(PubChem_ID) %>% t %>% as.character %>% unique 
# remove NAs
drugs_missing = drugs_missing[complete.cases(drugs_missing)]
write.table(drugs_missing, 'PubChem_ID_missing_metabolites.txt', quote = F, col.names = F, row.names = F)


# after downloading them from pubchem, the file is named as: Metab_structures_missing_biolog.sdf

# file with chem fingerprints from pubchem compounds (~100 more, not bad)
biolog_metabolites_PubChem = read_csv("biolog_metabolites_PubChem.csv") %>% select(-X1, -index, -Number_of_carboxylic_acids_1, -ID) %>%
  rename(PubChem_ID = PUBCHEM_COMPOUND_CID) 


dummy = biolog %>%
  filter(Metabolite %in% missing) %>%
  select(Metabolite, PubChem_ID) %>% unique

biolog_metabolites_PubChem = biolog_metabolites_PubChem %>% left_join(dummy, by = 'PubChem_ID') %>%
  select(PubChem_ID, Metabolite, everything()) %>% rename(GENERIC_NAME = Metabolite)


# bind rows
complete = bind_rows(drugbank, biolog_metabolites_PubChem) %>% 
  distinct(GENERIC_NAME, .keep_all = TRUE)

# save data table into csv
write.csv(complete, 'Complete_chem_fingerprints.csv')
```

With the new data, let's proceed with the analysis.

1. Do the PCA: 

```{r}
mat = complete %>% 
  filter(GENERIC_NAME != 'Tannic acid') %>%
  select(Number_of_aliphatic_carboxylic_acids:Number_of_urea_groups) %>% as.matrix
rownames(mat) = complete %>% filter(GENERIC_NAME != 'Tannic acid') %>% select(GENERIC_NAME) %>% t %>% as.character


res.pca = PCA((mat), scale.unit = TRUE, ncp = 5, graph = F)
ind = get_pca_ind(res.pca)
ind_df = data.frame(ind$coord[,1], ind$coord[,2], ind$coord[,3])
colnames(ind_df) = c('Dim1', 'Dim2', 'Dim3')
```


2. Plot the data:

```{r}
test2 = ind_df %>% tibble 
test2['Drugbank'] = complete %>% filter(GENERIC_NAME != 'Tannic acid') %>% select(GENERIC_NAME) %>% t %>% as.character

test2 = test2 %>%
  mutate(biolog = ifelse(Drugbank %in% drugs, Drugbank, NA),
         Category = ifelse(Drugbank %in% drugs, 'Biolog', 'Drugbank'),
         Category = as.factor(Category)) 


fig = plot_ly(data = test2,  x = ~Dim1, y = ~Dim2, z = ~Dim3, text = test2$Drugbank, 
               color = ~Category, colors = c('#FF0900', '#B8B8B8'),
               marker = list(size = 3))
fig
```


Notice that I was removing [**Tanic Acid**](https://pubchem.ncbi.nlm.nih.gov/compound/Tannic-acid) from the plot because it's a super different compound that gets extremely separated from the rest. In any case, there is a major difference between this plot and the previous one because the cloud of points have a different shape. I don't think this is very important, but we need to have this in mind. 



## t-SNE representation

PCA is a good method to reduce the dimensionality, but it's showing some weird results compared to the original publication. Before testing more things with the PCA parameters and transformations, let's give it a try with the t-SNE method, which in theory is more robust to show non-linear dependencies and better clusters than a PCA. 

```{r}

library(Rtsne)

mat.tsne = complete %>% 
  select(Number_of_aliphatic_carboxylic_acids:Number_of_urea_groups) %>% as.matrix
rownames(mat.tsne) = complete %>% 
  select(GENERIC_NAME) %>% t %>% as.character
```

This time we have not removed Tanic Acid, so let's see what happens with the clustering. First of all, let's run the tsne method. If you want to know more about the paramters, especially the `perplexity` parameter, go to this [website](https://distill.pub/2016/misread-tsne/) and play with the examples. 
The code might take a while. If you are running in a slower computer, reduce the `max_iter` parameter to 1000. 

```{r}

tsne = Rtsne(mat.tsne, check_duplicates = FALSE, pca = FALSE, num_threads = 12,
             normalize = FALSE, max_iter = 1000,
             perplexity = 40, theta = 0.2, dims = 3) 
```

And let's plot what we have doing the proper data transformations:

```{r warning=FALSE, message=FALSE}
# generate data frame from tnse results
tsne.df = data.frame(tsne$Y)
colnames(tsne.df) = c('Dim1', 'Dim2', 'Dim3')

tsne.df['Drugbank'] = complete %>% 
  # filter(GENERIC_NAME != 'Tannic acid') %>% 
  select(GENERIC_NAME) %>% t %>% as.character

tsne.df = tsne.df %>%
  mutate(biolog = ifelse(Drugbank %in% drugs, Drugbank, NA),
         Category = ifelse(Drugbank %in% drugs, 'Biolog', 'Drugbank'),
         Opacity = ifelse(Category == 'Biolog', 1, 0.1),
         Opacity = as.numeric(Opacity),
         Category = as.factor(Category)) 


fig = plot_ly(data = tsne.df,  x = ~Dim1, y = ~Dim2, z = ~Dim3, text = tsne.df$Drugbank, 
               color = ~Category, colors = c('#FF0900', '#B8B8B8'),
               marker = list(size = 4))
fig
```

Good! It's much nicer than the PCA in my opinion. But there's still the detail of the opacity for the grey points. I haven't got a proper way to make it work with values within the dataframe, but we can make a slightly more convoluted code to plot it in a nicer way.

```{r message=FALSE, warning=FALSE}
# let's try to modify the opacity of grey points
fig = plot_ly()

fig = fig %>%
  add_trace(
    data = tsne.df %>% filter(Category == 'Drugbank'),
    name = 'Drugbank',
    x = ~Dim1, y = ~Dim2, z = ~Dim3, text = tsne.df[tsne.df$Category == 'Drugbank',]$Drugbank, 
    marker = list(
      color = '#B8B8B8', 
      opacity = 0.4, # OPACITY
      size = 4)
    )

fig = fig %>%
  add_trace(
    data = tsne.df %>% filter(Category == 'Biolog'),
    name = 'Biolog',
    x = ~Dim1, y = ~Dim2, z = ~Dim3, text = tsne.df[tsne.df$Category == 'Biolog',]$Drugbank, 
    marker = list(
      color = '#FF0900', 
      size = 4)
  )
fig

```


## NEW! t-SNE Representation

After doing some work looking for IDs in Pubchem by hand, I have finally a list of 234 biolog drugs in the database. Some of them were repeated, and 3 of them are not as Pubchem IDs, but PubChem substances. I will check if I can make the Python script to work with those.

```{r warning=FALSE, echo=FALSE}
biolog.table = biolog %>% 
  filter(Plate %in% plates) %>% 
  select(Metabolite, Plate, KEGG_ID, CAS_ID, PubChem_ID) %>% 
  unique

# I LOOKED FOR THE MISSING COMPOUNDS AND ADDED THEM
# ALSO, I ADDED A NEW COLUMN WITH PUBCHEM SUBSTANCE
# AS 3 COMPOUNDS WERE NOT FOUND IN PUBCHEM

# load AGAIN the file, it has been updated
biolog_pubchem_IDs = read_csv("biolog_pubchem_IDs.csv")

# store the list of pubchem IDs
pubchem_IDs = biolog_pubchem_IDs %>% select(PubChem_ID) %>% t %>% as.character 

# store the list of pubchem substances
pubchem_subs = biolog_pubchem_IDs %>% select(PubChem_Substance) %>% t %>% as.character 


# file with chem fingerprints from pubchem compounds (~100 more, not bad)
biolog_metabolites_PubChem = read_csv("biolog_metabolites_PubChem.csv") %>% select(-X1, -index, -Number_of_carboxylic_acids_1, -ID) %>%
  rename(PubChem_ID = PUBCHEM_COMPOUND_CID) 

dummy = biolog_pubchem_IDs %>%
  select(Metabolite, PubChem_ID) %>% unique

biolog_metabolites_PubChem = biolog_metabolites_PubChem %>% left_join(dummy, by = 'PubChem_ID') %>%
  select(PubChem_ID, Metabolite, everything()) %>% rename(GENERIC_NAME = Metabolite)

# add metformin to the database
biolog_metabolites_PubChem[biolog_metabolites_PubChem$PubChem_ID == 4091,]$GENERIC_NAME <- 'Metformin'

# add a new category for biolog compounds
biolog_metabolites_PubChem = biolog_metabolites_PubChem %>% 
  mutate(Category = 'Biolog')

# bind rows
complete = bind_rows(drugbank %>% 
                       mutate(Category = 'Drugbank'), 
                     biolog_metabolites_PubChem)
```

```{r}
mat.tsne = complete %>% 
  # filter(GENERIC_NAME != 'Tannic acid') %>%
  # select(-Number_of_carboxylic_acids_1) %>%
  select(Number_of_aliphatic_carboxylic_acids:Number_of_urea_groups) %>% as.matrix
rownames(mat.tsne) = complete %>% 
  # filter(GENERIC_NAME != 'Tannic acid') %>% 
  select(GENERIC_NAME) %>% t %>% as.character

tsne = Rtsne(mat.tsne, check_duplicates = FALSE, pca = FALSE, num_threads = 12,
             normalize = FALSE, max_iter = 1000,
             perplexity = 40, theta = 0.2, dims = 3) 

```


And, plot!

```{r message=FALSE, warning=FALSE}
# generate data frame from tnse results
tsne.df = data.frame(tsne$Y)
colnames(tsne.df) = c('Dim1', 'Dim2', 'Dim3')

tsne.df['Drugbank'] = complete %>% 
  # filter(GENERIC_NAME != 'Tannic acid') %>% 
  select(GENERIC_NAME) %>% t %>% as.character


tsne.df = tsne.df %>%
  mutate(biolog = ifelse(Drugbank %in% drugs, Drugbank, NA),
         Category = ifelse(Drugbank %in% drugs, 'Biolog', 'Drugbank'),
         Category = ifelse(Drugbank == 'Metformin', 'Metformin', Category),
         Category = as.factor(Category)) 


# let's try to modify the opacity of grey points
fig = plot_ly() 

fig = fig %>%
  add_trace(
    data = tsne.df %>% filter(Category == 'Drugbank'),
    name = 'Drugbank',
    x = ~Dim1, y = ~Dim2, z = ~Dim3, text = tsne.df[tsne.df$Category == 'Drugbank',]$Drugbank, 
    marker = list(
      color = '#B8B8B8', 
      opacity = 0.4, # OPACITY
      size = 4)
  )

fig = fig %>%
  add_trace(
    data = tsne.df %>% filter(Category == 'Biolog'),
    name = 'Biolog',
    x = ~Dim1, y = ~Dim2, z = ~Dim3, text = tsne.df[tsne.df$Category == 'Biolog',]$Drugbank, 
    marker = list(
      color = '#FF0900', 
      size = 4)
  )

fig = fig %>%
  add_trace(
    data = tsne.df %>% filter(Category == 'Metformin'),
    name = 'Metformin',
    x = ~Dim1, y = ~Dim2, z = ~Dim3, text = tsne.df[tsne.df$Category == 'Metformin',]$Drugbank, 
    marker = list(
      color = '#3E4AFF', 
      size = 6)
  )


fig
```



## NEW!! Synergistic/antagonistic compounds

After doing the analysis of synergy scores with SynergyFinder, we can now colour-code the points accordingly to see if there are clusters. 


```{r message=FALSE, warning=FALSE, echo=FALSE}

bliss = read_csv('D:/MRC_Postdoc/Pangenomic/biolog/Biolog_metf_40_60/summary/Bliss_scores_corrected.csv')

bliss = bliss %>% 
  mutate(Drug.combination = substr(Drug.combination, 1, nchar(Drug.combination) - 6)) 

# initiate the threshold
thrs = 4

### BE CAREFUL AND CHOSE A SPECIFIC WAY TO SEPARATE COMPOUNDS
# classify by CI_low
bliss.tsne = bliss %>% 
  mutate(CI_low = abs(Synergy.score) - abs(CI),
         CI_up = abs(Synergy.score) + abs(CI),
         Direction = ifelse(CI_low > thrs & Synergy.score < 0, 'Antagonistic', 
                            ifelse(CI_low > thrs & Synergy.score > 0, 'Synergistic', 'Neutral')),
         Drug.combination = fct_reorder(Drug.combination, desc(Synergy.score))) %>%
  rename(Drugbank = Drug.combination)



tsne.df = tsne.df %>% left_join(bliss.tsne)
```


```{r message=TRUE, warning=FALSE}
# let's try to modify the opacity of grey points
fig = plot_ly() 

fig = fig %>%
  add_trace(
    data = tsne.df %>% filter(Category == 'Drugbank'),
    name = 'Drugbank',
    x = ~Dim1, y = ~Dim2, z = ~Dim3, text = tsne.df[tsne.df$Category == 'Drugbank',]$Drugbank, 
    marker = list(
      color = '#B8B8B8', 
      opacity = 0.4, # OPACITY
      size = 4)
  )

fig = fig %>%
  add_trace(
    data = tsne.df %>% filter(Direction == 'Antagonistic'),
    name = 'Antagonistic',
    x = ~Dim1, y = ~Dim2, z = ~Dim3, text = (tsne.df %>% filter(Direction == 'Antagonistic'))$Drugbank ,
    marker = list(
      color = '#FF0900', 
      size = 6)
  )

fig = fig %>%
  add_trace(
    data = tsne.df %>% filter(Direction == 'Synergistic'),
    name = 'Synergistic',
    x = ~Dim1, y = ~Dim2, z = ~Dim3, text = (tsne.df %>% filter(Direction == 'Synergistic'))$Drugbank, 
    marker = list(
      color = '#057D33', 
      size = 6)
  )

fig = fig %>%
  add_trace(
    data = tsne.df %>% filter(Direction == 'Neutral'),
    name = 'Neutral',
    x = ~Dim1, y = ~Dim2, z = ~Dim3, text = (tsne.df %>% filter(Direction == 'Neutral'))$Drugbank, 
    marker = list(
      color = '#4D5280', 
      size = 4,
      opacity = 0.7)
  )

fig = fig %>%
  add_trace(
    data = tsne.df %>% filter(Category == 'Metformin'),
    name = 'Metformin',
    x = ~Dim1, y = ~Dim2, z = ~Dim3, text = tsne.df[tsne.df$Category == 'Metformin',]$Drugbank, 
    marker = list(
      color = '#3E4AFF', 
      size = 6)
  )


fig

```


<br><br><br>








